Metric Spaces with Expensive Distances

### 2020-09-20 paper/ijcga-expensive-distances.tex
ijcga draft
Fix typos and small reviewers comments.

\documentclass{ws-ijcga}

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance between a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.

@@

\myparagraph{Well-Separated Pair Decomposition}
Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
if 
\[
\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
\]
A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ 
of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
Har-Peled and Mendel \cite{hm-fast} 
introduced the notion of net-trees and
generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
\begin{enumerate}
    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
    \item If $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time. The algorithm uses the net-tree structure.
\end{enumerate}
The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
than net-trees. We mention in passing that cover trees can also be used for building
a spanner (this can be proven with the same methods), and we use cover trees for
building WSPD spanners in one of our implementations.

@@

On the other hand, spanner constructions based on Well-Separated Pair Decomposition~\cite{cal-kos-wspd,hm-fast} only compute
$O(n\log n+n \eps^{-d})$ distances 
to construct a \espanner in doubling dimension $d$.
The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
computations. However, the algorithm is significantly more involved.

\section{Blind spanners}
\label{sec:blind_spanners}
%
We introduce a new framework for constructing spanners
which we call \emph{blind spanners}: the idea is to maintain,
for every pair of points $(p_i,p_j)$,
a lower bound $a_{ij}$ and an upper bound $b_{ij}$ for $\dist(p_i,p_j)$,
initially set to $0$ and $\infty$, respectively. While there exists some pair for which $\frac{b_{ij}}{a_{ij}}>(1+\eps)$,
we pick one of them, compute its distance and update the lower and upper bounds of
all pairs with respect to the newly acquired information. Here is the pseudocode:

@@

We performed three types of experiments: on points sampled from a low-dimensional Euclidean
space, on images with Earth Mover's distance, and on persistence diagrams
with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See Section~\ref{sec:appendix_background} for the definition of $\wsdist{q}$.}
Clearly, our model is not meaningful in Euclidean spaces because
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

From our experiments, we conclude that the algorithm with the highest chance
of saving a significant amount of distance computations on real data is \bgrdy. 
Even though \brndm also substantially reduces
the number of computed distances, its performance is worse;
other algorithms that we tested do not produce sparse spanners.
While it would seem plausible that access to approximate value of the distance
could be exploited in the spanner construction,
we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
dense spanners. More data from experiments in $\RR^n$ can be found in
Section~\ref{sec:spanners_euclidean_addendum}.

@@

Alternatively, we also experimented with the Earth Mover’s distance on
gray value images by considering each image as a distribution on pixels.
For this (non-standard) variant, we obtain even larger speed-ups~-- see
Section~\ref{sec:spanners_euclidean_addendum} for details.

@@

The full proof is given in Section~\ref{sec:proof_ann},
we summarize the main ideas:
first, we notice that points ``far away'' from the current candidate $c$ 
can be disregarded in the algorithm. More precisely, if $p_i$
is outside of the ball of radius $2\dist(c, q)$ centered at $c$,
the lower bound for $p_i$ will be so large that
the distance to of $p_i$ to $q$ does not have to be queried.
Second, whenever a distance of $q$ to some $p_j$ is computed,
we can disregard all points ``very close'' to $p_j$.
Precisely, points in a ball of radius $\eps\dist(p_j,q)$
will not invalidate the current candidate 
to be an approximate nearest neighbor and hence,
no distance query for such points is needed.

@@

The proof strategy fails for $\eps=0$ because the $\eps$-ball
around $p_j$ as described above degenerates, and the packing argument fails.
Indeed, as the example in Figure~\ref{fig:exact_bad_example} (in Section~\ref{sec:proof_ann}) 
shows, there are point sets where the expected number
of distance computations for exact nearest neighbor is linear.

@@

The existence of a $2$-approximation algorithm did not help us to significantly
reduce the number of exact distance computations, although it seems obvious
that knowing the all approximate distances is useful.
We pose the question what heuristic could make more use of this feature.

@@

\begin{proof} (of Theorem~\ref{thm:ann_bound})
%
In the sequence $p_1,\ldots,p_n$, let $p_k$ be a point 
such that $r_i<r_k$ for all $1\leq i\leq k-1$. We call an element
of this form a \emph{minimum} of the sequence. 
A standard backwards analysis argument~\cite{seidel-backwards} shows that the probability
of $p_k$ being a minimum is at most $1/k$, so that the number of minima
in the sequence is $O(\log n)$ in expectation.

@@

\myparagraph{Wasserstein Distance}
Let $\diag$ denote the line in the plane with the equation $y = x$; we call it the diagonal.
A persistence diagram is a finite multiset of points $(x, y)$ with $x > y$
$\Delta$ and all points of $\Delta$ with infinite cardinality.
The points with $y > x$ are called \textit{off-diagonal} points,
and points on $\Delta$ are called \textit{diagonal} points.
The \emph{$q$-Wasserstein distance} is defined as
\[
    \wsdist{q}(X,Y) = \left[ \inf_{\eta:X\rightarrow Y} \sum_{x\in X} \|x-\eta(x)\|_\infty^q\right]^{1/q},
\]
where $\eta$ ranges over all bijections between $X$ and $Y$.
Equivalent definition, which is better suited for actual computations,
is as follows. Let $\pi\colon \RR^2 \to \Delta$ be defined
as $(x,y) \mapsto (\frac{x+y}{2}, \frac{x+y}{2})$; we refer to $\pi$ as \textit{diagonal
projection}. Let $\tX$ be the set $(X\setminus \Delta) \cup \pi(Y \setminus \Delta)$
and $\tY$ be the set $(Y\setminus \Delta) \cup \pi(X \setminus \Delta)$; in plain words,
to obtain $\tX$, we add to the off-diagonal points of $X$
the diagonal projections of the off-diagonal points of $Y$,
and perform the same operation to obtain $\tY$. Notice that $\tX$ and $\tY$
are finite sets of the same cardinality. Let $G$ be a complete bipartite
graph with vertices $\tX \coprod \tY$, where each vertex from $\tX$
is connected with each vertex from $\tY$. If edge $(x,y)$ of $G$
has at least one of the vertices $x$ or $y$ not on the diagonal,
then the edge is assigned weight $\ell_{\infty}(x, y)^q$, otherwise
the edge is assigned weight 0. Wasserstein distance is defined as the $q$-th
root of the minimal cost of a perfect matching on $G$; in other words,
in order to find $\wsdist{q}$, one must solve an \textit{assignment problem}.

>>>>

\documentclass[]{ws-ijcga}

@@

We propose several heuristics to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computations. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance between a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.

@@

\myparagraph{Well-Separated Pair Decomposition}
Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
if 
\[
\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
\]
A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ 
of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
Har-Peled and Mendel \cite{hm-fast} 
introduced the notion of net-trees and
generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
\begin{enumerate}
    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
    \item If $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time. The algorithm uses the net-tree structure.
\end{enumerate}
The algorithm for constructing a net-tree is complicated and not easy to implement. Beygelzimer
et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
than net-trees. We mention in passing that cover trees can also be used for building
a spanner (this can be proven with the same methods), and we use cover trees for
building WSPD spanners in one of our implementations.

@@

On the other hand, spanner constructions based on Well-Separated Pair Decomposition~\cite{cal-kos-wspd,hm-fast} only compute
$O(n\log n+n \eps^{-d})$ distances 
to construct a {\espanner} in doubling dimension $d$.
The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
computations. However, the algorithm is significantly more involved.

\section{Blind spanners}
\label{sec:blind_spanners}
%
We introduce a new framework for constructing spanners
that we call \emph{blind spanners}: the idea is to maintain,
for every pair of points $(p_i,p_j)$,
a lower bound $a_{ij}$ and an upper bound $b_{ij}$ for $\dist(p_i,p_j)$,
initially set to $0$ and $\infty$, respectively. While there exists some pair for which $\frac{b_{ij}}{a_{ij}}>(1+\eps)$,
we pick one of them, compute its distance and update the lower and upper bounds of
all pairs with respect to the newly acquired information. Here is the pseudocode:

@@

We performed three types of experiments: on points sampled from a low-dimensional Euclidean
space, on images with Earth Mover's distance, and on persistence diagrams
with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \ref{sec:appendix_background} for the definition of $\wsdist{q}$.}
Clearly, our model is not meaningful in Euclidean spaces because
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

From our experiments, we conclude that the algorithm with the highest chance
of saving a significant amount of distance computations on real data is \bgrdy. 
Even though \brndm also substantially reduces
the number of computed distances, its performance is worse;
other algorithms that we tested do not produce sparse spanners.
While it would seem plausible that access to approximate value of the distance
could be exploited in the spanner construction,
we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
dense spanners. More data from experiments in $\RR^n$ can be found in
\ref{sec:spanners_euclidean_addendum}.

@@

Alternatively, we also experimented with the Earth Mover’s distance on
gray value images by considering each image as a distribution on pixels.
For this (non-standard) variant, we obtain even larger speed-ups~--- see
\ref{sec:spanners_euclidean_addendum} for details.

@@

The full proof is given in \ref{sec:proof_ann},
we summarize the main ideas:
first, we notice that points ``far away'' from the current candidate $c$ 
can be disregarded in the algorithm. More precisely, if $p_i$
is outside of the ball of radius $2\dist(c, q)$ centered at $c$,
the lower bound for $p_i$ will be so large that
the distance to of $p_i$ to $q$ does not have to be queried.
Second, whenever a distance of $q$ to some $p_j$ is computed,
we can disregard all points ``very close'' to $p_j$.
Precisely, points in a ball of radius $\eps\dist(p_j,q)$
will not invalidate the current candidate 
to be an approximate nearest neighbor and hence,
no distance query for such points is needed.

@@

The proof strategy fails for $\eps=0$ because the $\eps$-ball
around $p_j$ as described above degenerates, and the packing argument fails.
Indeed, as the example in Figure~\ref{fig:exact_bad_example} (in \ref{sec:proof_ann}) 
shows, there are point sets where the expected number
of distance computations for exact nearest neighbor is linear.

@@

The existence of a $2$-approximation algorithm did not help in significantly
reducing the number of exact distance computations, although it seems obvious
that knowing the all approximate distances is useful.
We pose the question what heuristic could make more use of this feature.

@@

\begin{proof} (of Theorem~\ref{thm:ann_bound})
%
In the sequence $p_1,\ldots,p_n$, let $p_k$ be a point 
such that $r_i > r_k$ for all $1\leq i\leq k-1$. We call an element
of this form a \emph{minimum} of the sequence. 
A standard backwards analysis argument~\cite{seidel-backwards} shows that the probability
of $p_k$ being a minimum is at most $1/k$, so that the number of minima
in the sequence is $O(\log n)$ in expectation.

@@

\myparagraph{Wasserstein Distance}
Let $\diag$ denote the line in the plane with the equation $y = x$; we call it the diagonal.
A \textit{persistence diagram} is a
multiset that consists of all points of $\diag$ with infinite multiplicity
and finitely many points $(x, y)$ with $y > x$, each with finite multiplicity.
The points with $y > x$ are called \textit{off-diagonal} points,
and the points on $\diag$ are called \textit{diagonal} points.
Let $X$ and $Y$ be two persistence diagrams.
The \emph{$q$-Wasserstein distance} between $X$ and $Y$ is defined as
\[
    \wsdist{q}(X,Y) = \left[ \inf_{\eta:X\rightarrow Y} \sum_{x\in X} \|x-\eta(x)\|_\infty^q\right]^{1/q},
\]
where $\eta$ ranges over all bijections between $X$ and $Y$.
Equivalent definition, which is better suited for actual computations,
is as follows. Let $\pi\colon \RR^2 \to \diag$ be defined
as $(x,y) \mapsto (\frac{x+y}{2}, \frac{x+y}{2})$; we refer to $\pi$ as \textit{diagonal
projection}. Let $\tX$ be the set $(X\setminus \diag) \cup \pi(Y \setminus \diag)$
and $\tY$ be the set $(Y\setminus \diag) \cup \pi(X \setminus \diag)$; in plain words,
to obtain $\tX$, we add to the off-diagonal points of $X$
the diagonal projections of the off-diagonal points of $Y$,
and perform the same operation to obtain $\tY$. Notice that $\tX$ and $\tY$
are finite sets of the same cardinality. Let $G$ be a complete bipartite
graph with vertices $\tX \coprod \tY$, where each vertex from $\tX$
is connected with each vertex from $\tY$. If edge $(x,y)$ of $G$
has at least one of the vertices $x$ or $y$ not on the diagonal,
then the edge is assigned weight $\ell_{\infty}(x, y)^q$, otherwise
the edge is assigned weight 0. Wasserstein distance is defined as the $q$-th
root of the minimal cost of a perfect matching on $G$; in other words,
in order to find $\wsdist{q}$, one must solve an \textit{assignment problem}.
